---
title: "Characterization of missing values in untargeted MS-based metabolomics data and evaluation of missing data handling strategies"
subtitle: "Example code for data simulation and imputation "
author: "K.T. Do\\*, S. Wahl\\*, J. Raffler, S. Molnos, M. Laimighofer, J. Adamski, K. Suhre, K. Strauch, A. Peters, C. Gieger, C. Langenberg, I.D. Stewart, F.J. Theis, H. Grallert, G. Kastenmueller\\#, J. Krumsiek\\#"
date: "27.06.2018"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document:
    toc: FALSE
    number_sections: FALSE
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=7, 
                      echo=T, warning=FALSE, message=FALSE, results="show", eval=T)
```


# Simulation of incomplete data


## Load required packages and functions

```{r eval=FALSE}
require(mvtnorm)
require(parallel)
require(corpcor)

source("gen.data.met.r")
```


## Parameter settings
```{r eval=FALSE}
n <- 250
naux <- 5	
p <- 2+2*naux
corr <- 0.2
pcor12 <- 0.09977889 # corresponding to Pearson correlation corr = 0.2
pcoraux <- 0.3165772
```

- `n` is the number of observations.
- `naux` is the number of auxiliary metabolites for each metabolite of interest.
- `p` is the number of variables to be created; in this case, 2 metabolites of interest and 5 auxiliary metabolites for each of them are created.
- `corr` is the Pearson correlation to achieve between the 2 metabolites of interest.
- `pcor12` is the partial correlation between the 2 metabolites of interest such that the set Pearson correlation ($corr=0.2$) is achieved in the overall correlation matrix. 
- `pcoraux` is the partial correlation of each metabolite of interest with its 5 auxiliary metabolites. In this case, the value was determined numerically based on our real dataset (KORA F4) such that the 5 auxiliary metabolites achieve an adjusted $R^2$ of 50% for the metabolite of interest on average to resemble real data example.


## Generate correlation matrix based on specified parameters

```{r eval=FALSE}
Pcor=rbind(c(0,pcor12,rep(pcoraux,naux),rep(0,naux)), 
           c(pcor12,0,rep(0,naux),rep(pcoraux,naux)),
          matrix(rep(c(pcoraux,0,rep(0,naux),rep(0,naux)),each=naux),nrow=naux),
          matrix(rep(c(0,pcoraux,rep(0,naux),rep(0,naux)),each=naux),nrow=naux))
diag(Pcor) <- 1

Cor <- pcor2cor(Pcor) 
```

- `Pcor` is the generated partial correlation matrix.
- `Cor` is the Pearson correlation matrix generated from the partial correlation matrix using function `pcor2cor` from package `corpcor`.


## Generate artificial, complete dataset

Based on the parameters defined above, the function `gen.complete.data` generates a list `datCom`, which stores the artificial data in `datCom$x` and runday information in `datCom$rundays`. `datCom$x` is a data.frame with 250 observations and 13 variables. The first and second column correspond to the two main metabolites of interest, columns 3-12 correspond to the auxiliary metabolites, and the last column stores the phenotype (see Supporting Information S3).

```{r eval=FALSE}
datCom <- gen.complete.data(n=n,
							p=p,
							Cor=Cor,
							n.rundays=ceiling(n/34))
```

- `n` is the number of observations.
- `p` is the number of variables.
- `Cor` is correlation matrix of the variables.
- `n.rundays` is the number of rundays. In this case, it is specified such that each runday comprises measurements of 34 samples, the everage number observed for the KORA F4 dataset.


## Generate artificial, incomplete data by introducing missingness

The function `gen.miss` introduces missing values into the generated artificial data matrix, thereby producing the incomplete data matrix `dat`.

```{r eval=FALSE}
dat <- gen.miss(datx=datCom$x,
				incoms=c(0.1,0.1,rep(0,naux*2+1)),
				mech="trend.runday",
				betas1 = rep(-10,2),
        strengths.incom.vary=rep(0.5/4^2,2),
				strengths.beta1.vary=rep(25,2),
        cors.incom=rep(0,2),
				rundays=datCom$rundays)

dat <-  data.frame(dat,rundays=datCom$rundays)
datCom <- data.frame(datCom$x,rundays=datCom$rundays)
```

- `datx` is the complete data set (output from `gen.complete.data`).
- `incoms` is the vector of the proportion (percentage) of incomplete entries to be introduced into the variables; in this example, 10% of the values of the two main metabolites will be set to missing, while all remaining variables will remain complete.
- `mech` is the missingness mechanism to be introduced into the data (see Supporting Information S3); in this case, runday-specific probabilistic LOD is used.
- `betas1` is a vector which specifies the strength of dependency of missingness from the concentration values; in this case, we specify a strong, negative dependency guided by observations from the real dataset (see parameter $\beta_1$ in Table S8 of Supporting Information S3).
- `strengths.incom.vary` is a vector which specifies the variation of missingness across rundays; in this case, we set $(min(incom,1-incom)/4)^2$ for moderate variation, which was guided by observations from the real data (see parameter $\sigma^2_{miss}$ in Table S8 of Supporting Information S3).
- `strengths.beta1.vary` is a vector which specifies the variation of beta1 across rundays; in this case, we set it to $(\beta_1/2)^2$ for strong variation, which was guided by observations from the real data (see parameter $\sigma^2_{\beta_1}$ in Table S8 of Supporting Information S3).
- `cors.incom` specifies the correlation of missingness across rundays for the two main metabolites (see parameter $\rho_{miss}$ in Table S8 of Supporting Information S3).
- `rundays` defines the number of rundays in which each sample was measured.




# Imputation and regression- and correlation-based analysis

## Load required packages and functions

```{r eval=FALSE}
require(mice)
require(mitools)
require(pan)
require(ppcor)
require(tmvtnorm)
source("do.imp.met.r")
```


## Perform imputation and evaluations

The function `do.imp` performs imputation for a given list of methods. The output can either be the imputed data or the already evaluated performance results of the different imputation methods.

```{r eval=FALSE}
results <- do.imp(dat,
				  datCom,   
				  methods=c("complete", "cca", "min","mean","nie","nie.runday", "u.tsi",
				            "mi.avg.norm", "mi.avg.pmm", "u.tsi.runday","si.norm","si.pmm", 
				            "u.tsmi","u.tsmi.runday", "mi.norm","mi.pmm", "si.pan.incl.runday",
				            "mi.pan.incl.runday", "knn.variable","knn.sample.euc",
				            "knn.sample.euc.sel"),
				  #analysis = c("mse1","mse2","mse3","mse4","corxy","pcorxy","lmx","lmy","logx"),
				  K=c(1,3,5,10,20),
				  min.in.runday=17,
				  return.da1=T)
```

- `dat` is the incomplete data set generated by the function `gen.miss`; the two columns correspond to the two main metabolites, the third column contains a continuous phenotype without missing values.
- `datCom` (optional) is the corresponding complete data set generated by the function `gen.complete.data` for comparison of the results.

- `methods` is a vector of imputation methods to be applied (see Supporting Information S4).
    * cca: complete case analyses
    * min: minimum imputation
    * mean: mean imputation
    * nie: Richardson & Ciampi (RC)
    * nie.runday: RC on rundays (RC-R)
    * u.tsi: truncated sampling imputation (ITS)
    * u.tsi.runday: ITS on rundays (ITS-R)
    * si.norm: multivariate single imputation by chained eqations and Bayesian regression as primary model (ICE-norm)
    * si.pmm: ICE and predictive mean matching as primiary model (ICE-pmm)
    * u.tsmi: multiple truncated sampling imputation (MITS)
    * u.tsmi.runday: MITS on rundays (MITS-R)
    * mi.norm: multiple ICE-norm (MICE-norm)
    * mi.pmm: multiple ICE-pmm (MICE-pmm)
    * si.pan.incl.runday: ICE with random intercept per runday (ICE-adjR)
    * mi.pan.incl.runday: MICE with random intercept per runday (MICE-adjR)
    * knn.variable: k-nearest neighbor imputation on variables (KNN-var)
    * knn.sample.euc: k-nearest neighbor imputation on samples based on Euclidian distance (KNN-obs)
    * knn.sample.euc.sel: KNN-obs with variable pre-selection
    
- `analysis` (optional) is a vector of pervormance evalation methods to be performed.
    * mse1/2/3/4: mean squared error
    * corxy: correlation between the two main metabolites
    * pcorxy: partial correlation between the two main metabolites
    * lmx: linear regression with first metabolite as predictor and second metabolite as response
    * lmy: linear regression with first metabolite as response and second metabolite as predictor
    * logx: logistic regression with phenotype as response and first metabolite as predictor
    
- `K` is a vector of neighbors to use for k-nearest neighbor imputation
- `min.in.runday` specifies the minimum number of observed values in each runday used as a condition to perform ITS imputation; in this example, we set it to half of the average number of observations per runday observed in the real data.
- `return.da1` is a logical parameter indicating that the imputed data should be returned (instead of the evaluation results).
